% Use draft for fast compile while editing; remove for final PDF
\documentclass[draft,sigplan,10pt,nonacm]{acmart}

\AtBeginDocument{%
\providecommand\BibTeX{{Bib\TeX}}}

\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric}

\title{Recovering TASO Substitutions Under Bounded Rewrite Closure}

\author{Pietro Marsella}
\email{marsella@stanford.edu}

\author{Bodo Wirth}
\email{bodow@stanford.edu}

\renewcommand{\shortauthors}{Marsella and Wirth}

% TODO: be consistent: always call them substitutions, not rewrite rules.

\begin{document}

\begin{abstract}
Machine learning compilers have emerged in recent years as a major tool for automatically optimizing the performance of deep learning models. A major component of this optimization pipeline is \textbf{algebraic rewriting}. In algebraic rewriting, given a set of substitutions, the compiler repeatedly applies these substitutions to the model in order to minimize some cost metric, such as runtime.

In several modern compilers, these substitutions are hard-coded. However, more recent approaches such as TASO aim to automatically discover them from a starting set of base axioms. In this paper, we study the following question: given a sufficiently expressive rewrite system, is it possible to recover, as compositions of the base TASO axioms, the substitutions discovered by TASO?

To answer this question, we implement a bidirectional rewrite system over a symbolic graph IR. Our system is able to recover over 95\% of the TASO substitutions, suggesting that most of them can be derived from a comparatively small axiom basis rather than treated as independent primitive rules.
\end{abstract}
\maketitle

\section{Introduction}
Machine learning compilers have become a central part of the deep learning software stack. Their role is not only to lower models to kernels, but also to transform computation graphs into equivalent forms that execute more efficiently on the target hardware. At this level, one of the most important optimization mechanisms is algebraic rewriting: replacing a subgraph with a semantically equivalent one that yields lower cost under some objective, such as runtime, memory traffic, or communication volume. As deep learning systems have grown larger, more heterogeneous, and more distributed, the quality of these graph-level transformations has become increasingly important.

A traditional way to build such optimizers is to hand-design a collection of graph substitutions. This requires substantial manual effort, scales poorly as the operator set grows, and makes it difficult to understand how much of the optimizer's behavior is really grounded in a small trusted semantic core. TASO addressed part of this problem by showing that useful graph substitutions can be generated automatically from a base set of operator axioms, then formally verified before being used in optimization~\cite{jia2019taso}. For the optimization step, TASO keeps a working set of equivalent graphs, sorted in order of cost, and greedily applies substitutions to the lowest cost graph for a predefined number of steps.

Subsequent work focused largely on how to search more effectively over the space induced by such substitutions. MetaFlow observed that restricting attention to immediately profitable rewrites can be too limiting, and introduced relaxed graph substitutions that allow performance-decreasing intermediate steps in order to reach better final graphs~\cite{jia2019relaxed}. TENSAT instead performs equality saturation to compactly represent the equivalent graphs at once and thereby alleviate the phase-ordering problems of sequential rewriting~\cite{yang2021equality,willsey2021egg}. Unity extended the same perspective to distributed training, representing both algebraic transformations and parallelization choices as substitutions over a unified parallel computation graph~\cite{unger2022unity}. The effectveiness of these systems suggests that having a general framework for gneerating and applying substitutions outperforms hand-designed ones.

This paper studies a complementary question. Rather than asking how to discover substitutions, or how to search over them once they are known, we ask whether, for some sufficiently powerful rewrite system, the substitutions discovered by TASO are already implicit in the bounded rewrite closure of TASO's own base axioms. This question matters for both methodology and engineering. If most discovered substitutions are derivable, then they can be viewed as lemmas induced by a comparatively small trusted core, rather than as independent primitive facts. If they are not derivable, then the gap points to limitations in the rewrite framework used to explore them.

To investigate this question, we implement a bidirectional rewrite system over a symbolic graph IR and use it to reconstruct TASO substitutions under bounded rewrite closure. Our investigation leads to a rewrite system that is able to handle several adversarial cases, including surjective variable matching, tensor aliasing, dead node elimination, and multi-output matching. Using this system, we recover over 95\% of the substitutions discovered by TASO. Beyond the headline number, the reconstruction process also provides a structural view of the substitution set itself: many substitutions emerge naturally as short derived lemmas, some require unexpectedly long derivations, and a small residual set appears to lie outside the closure induced by the current axiom basis.
\section{Motivation, Background, and Related Work}

% Cite TASO, Tensat, Unity papers.
% TASO operates as follows:
% 1. Establish a base set of operators.
% 2. Hand-design a set of axioms for these operators, expressed as first-order logic constraints.
% 3. Randomly generate the set of all graphs obtainable by composing these operators up to a chosen size
% bound. Call this set G.
% 4. Fuzz each graph and partition G into equivalence classes according to a fingerprint of the graph’s
% output behavior.
% 5. For each pair of graphs in the same equivalence class, use an SMT solver to formally prove (from the
% axioms) that substituting one graph for the other is sound.
% The resulting proven substitutions can then be plugged into any graph-substitution engine and used to optimize a given input graph.

Showing that TASO subsittuiosn are recoverable directly as a rewrite composition of the base axioms would significnatly simplify the compiler pipeline: rather than a separate subsystem to discover these substitutions and then a graph rewrite engine to apply them, one could simply directly use the rewrite system and the axioms instead.

TODO (e.g., TASO~\cite{jia2019taso})

\section{Approach and Implementation}
\subsection{IR}

The core data structure of the IR is the `Graph` type, which is a set of bindings between tensor variables and expressions, expressed an SSA-style term graph.
Each expression represents some operator which takes in both tensor variables and non-tensor terms (such as scalars, strides, ...) as inputs, and produces a single tensor as output.
The non tensor terms can be both variables or specific concrete values.
`Input` and `Output` are special operators, which denote which tensor vairables are inputs and outputs of the graph.

The following is a possible graph expressible in the IR:

\begin{verbatim}
x = Input()
y = Transpose(x)
z = Mul(y, alpha)
w = MatMul(y, z)
o = Output(w)
\end{verbatim}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  >=Latex,
  node distance=5mm and 10mm,
  tensor/.style={draw, circle, minimum size=8mm, inner sep=0pt},
  op/.style={draw, rounded corners, minimum width=16mm, minimum height=7mm, align=center},
  term/.style={draw, ellipse, minimum width=11mm, minimum height=7mm, inner sep=1pt}
]
\node[tensor] (x) {$x$};
\node[op, below=of x] (transpose) {Transpose};
\node[tensor, below=of transpose] (y) {$y$};

\node[op, below left=15mm and 13mm of y] (mul) {Mul};
% \node[term, left=of mul] (alpha) {$\alpha$};
\node[tensor, below=of mul] (z) {$z$};

\node[op, below right=15mm and 13mm of y] (matmul) {MatMul};
\node[tensor, below=of matmul] (w) {$w$};

\node[op, below=of w] (output) {Output};
\node[tensor, below=of output] (o) {$o$};

\draw[->] (x) -- (transpose);
\draw[->] (transpose) -- (y);
\draw[->] (y) -- (mul);
% \draw[->] (alpha) -- (mul);
\draw[->] (mul) -- (z);
\draw[->] (y) -- (matmul);
\draw[->] (z) -- (matmul);
\draw[->] (matmul) -- (w);
\draw[->] (w) -- (output);
\draw[->] (output) -- (o);
\end{tikzpicture}
\label{fig:ir-example}
\end{figure}

\subsection{Axioms}

Axioms are encoded as substitutions between two graphs: a source graph (the LHS of the substitution) and a destination graph (the RHS of the substitution).
This representation alone is not sufficient, because we must also specify how boundary tensors (the inputs and outputs of the graph) correspond across the two sides.
Each axiom therefore additionally stores a bijection between source and destination inputs, and a bijection between source and destination outputs.

For example, consider the substitution
\[
\forall x,y,z.\;
\mathsf{EwAdd}(\mathsf{MatMul}(x,y), \mathsf{MatMul}(x,z))
=
\mathsf{MatMul}(x,\mathsf{EwAdd}(y,z)).
\]
Here $x,y,z$ are input tensors, \texttt{MatMul} denotes matrix multiplication, and \texttt{EwAdd} denotes element-wise addition.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  >=Latex,
  tensor/.style={draw, circle, minimum size=8mm, inner sep=0pt},
  op/.style={draw, rounded corners, minimum width=18mm, minimum height=8mm, align=center}
]
\node at (-33mm, 22mm) {\textbf{src}};
\node[tensor] (xL) at (-48mm, 12mm) {$x$};
\node[tensor] (yL) at (-34mm, 12mm) {$y$};
\node[tensor] (zL) at (-18mm, 12mm) {$z$};
\node[op] (mm1L) at (-42mm, -2mm) {MatMul};
\node[op] (mm2L) at (-24mm, -2mm) {MatMul};
\node[op] (addL) at (-33mm, -17mm) {EwAdd};
\node[tensor] (outL) at (-33mm, -31mm) {$out$};
\draw[->] (xL) -- (mm1L);
\draw[->] (yL) -- (mm1L);
\draw[->] (xL) -- (mm2L);
\draw[->] (zL) -- (mm2L);
\draw[->] (mm1L) -- (addL);
\draw[->] (mm2L) -- (addL);
\draw[->] (addL) -- (outL);

\node at (0mm, -6mm) {$\Longrightarrow$};

\node at (33mm, 22mm) {\textbf{dst}};
\node[tensor] (xR) at (20mm, 12mm) {$x$};
\node[tensor] (yR) at (34mm, 12mm) {$y$};
\node[tensor] (zR) at (48mm, 12mm) {$z$};
\node[op] (addR) at (41mm, -2mm) {EwAdd};
\node[op] (mmR) at (28mm, -17mm) {MatMul};
\node[tensor] (outR) at (28mm, -31mm) {$out$};
\draw[->] (yR) -- (addR);
\draw[->] (zR) -- (addR);
\draw[->] (xR) -- (mmR);
\draw[->] (addR) -- (mmR);
\draw[->] (mmR) -- (outR);
\end{tikzpicture}
\label{fig:axiom-matmul-factor}
\end{figure}

\subsection{Substitution Matching}

Graph rewriting has two components: matching and application.
In the matching phase, we are given a target graph and a substitution, and we enumerate all valid occurrences of the substitution source inside the target.
Each successful match is represented as a surjective mapping from tensors in the substitution source graph to tensors in a target subgraph.

Matching starts by pairing an output tensor of the substitution source with a candidate tensor in the target graph.
We then recursively expand this pair under the following rules:
\begin{itemize}
\item If the source tensor is an input tensor, matching succeeds immediately (it may bind to either an input or an internal tensor in the target).
\item If the source tensor is internal but the target tensor is an input tensor, matching fails.
\item If both tensors are internal, their producing expressions must agree (same operator and compatible arguments). If they do not agree, matching fails.
\item If the expressions agree, we recursively match each corresponding pair of operand tensors.
\end{itemize}
During this recursion we maintain and extend a partial match mapping, enforcing consistency whenever a source tensor is revisited.
If recursion terminates without conflict, the resulting mapping is a valid match.

\subsection{Substitution Application}

%TODO: explain the application system

The rewrite system must be expressive enough to derive substitutions that TASO proves through SMT solving.

Our application system therefore handles the following cases:
\begin{itemize}
\item \textbf{Surjective variable matching.} For example, \texttt{matmul(x, y)} in the source pattern should match \texttt{matmul(x, x)}.
\item \textbf{Occurrence-based rewriting.} For example, given \texttt{matmul(x, x)}, the system should be able to rewrite to \texttt{matmul(x, x.T.T)} (i.e., only one occurrence of \texttt{x} is rewritten, not the entire node).
\item \textbf{Input/output aliasing.} Substitutions where input and output are aliased must be handled correctly. For example, \texttt{x.T.T -> x} should apply correctly even when aliasing occurs.
\item \textbf{Subgraph-local matching.} A pattern must be able to match any subgraph. For example, \texttt{x.T.T} should be applicable even when \texttt{x} is not a graph input and \texttt{x.T.T} is not a graph output.
\item \textbf{Dangling-expression cleanup.} The system must handle dangling expressions cleanly after rewriting.
\end{itemize}

The application system is implemented as follows:
%TODO add

\subsection{Search}

Our objective in search is the following: for a given TASO substitution, find a sequence of axiom applications that transforms the source graph of the substitution into a graph that is structurally equivalent to its destination graph.

A naive approach is to run BFS from the source graph while applying all axioms until the destination graph is reached. However, this is infeasible in practice because the number of reachable graphs grows exponentially with the search depth.

With the exception of two split axioms (\texttt{split0(concat(x,y)) -> x} and \texttt{split1(concat(x,y)) -> y}, which remove tensors), all axioms are invertible. This enables a more efficient strategy: bidirectional BFS. We run one search forward from the source using all axioms, and one search backward from the destination using inverse applications of invertible axioms. The search terminates when we find a graph reachable from both directions.

\subsection{Graph Canonicalization}

During search, we must quickly determine whether two graphs are equivalent. To do so, we apply the following canonicalization steps:
\begin{itemize}
\item Remove common subexpressions.
\item Remove unused tensors (except inputs and outputs, which must always be preserved).
\item Canonically rename tensor variables so that structurally equivalent graphs receive identical names.
\end{itemize}

Note that term graphs have ordered incoming edges, which makes isomorphism-style checks easier than for arbitrary graphs.

\section{Results}

Overal, we are able to recover XXX/576 of all the TASO substitutions.

% TODO: add table where we show how many steps is take to get from one graph to the other, and some other empirical results.

\section{Discussion, Limitations, and Next Steps}

The major surprise was

\subsection{Case Study}

\paragraph{Transpose(I) $\rightarrow$ I.}
Surprisingly, this test was initially failing due to an overly tight search budget. It turns out that this substitution, while derivable under the current axiom set, is more complicated than initially anticipated.

\noindent\textbf{Axioms used for the derivation.}
\begin{itemize}
\item \texttt{axiom26: matmul(x, I) -> x}
\item \texttt{axiom9: transpose(transpose(x)) -> x}
\item \texttt{axiom16: transpose(matmul(x, y)) ->}\\
\texttt{\hspace*{1.5em}matmul(transpose(y), transpose(x))}
\end{itemize}

\noindent In the traces below, \texttt{inv(axiom k)} means that \texttt{axiom k} is applied in the reverse direction.

\noindent\textbf{Proof trace for \texttt{transpose(I) -> I}.}
\begin{quote}
\footnotesize\ttfamily
(0) transpose(I) [start]\\
(1) matmul(transpose(I), I) [inv(axiom26)]\\
(2) matmul(transpose(I),\\
\hspace*{1.8em}transpose(transpose(I))) [inv(axiom9)]\\
(3) transpose(matmul(transpose(I),\\
\hspace*{1.8em}I)) [inv(axiom16)]\\
(4) transpose(transpose(I)) [axiom26]\\
(5) I [axiom9]
\end{quote}

\noindent\textbf{Analogous trace for \texttt{matmul(I, x) -> x}.}
\begin{quote}
\footnotesize\ttfamily
(0) matmul(I, x) [start]\\
(1) matmul(transpose(transpose(I)),\\
\hspace*{1.8em}x) [inv(axiom9)]\\
(2) matmul(transpose(transpose(I)),\\
\hspace*{1.8em}transpose(transpose(x))) [inv(axiom9)]\\
(3) transpose(matmul(transpose(x),\\
\hspace*{1.8em}transpose(I))) [inv(axiom16)]\\
(4) transpose(matmul(transpose(x), I)) [lemma: transpose(I) -> I]\\
(5) transpose(transpose(x)) [axiom26]\\
(6) x [axiom9]
\end{quote}

Note that the second derivation uses \texttt{transpose(I) -> I} as a lemma, making even the shortest proof several rewrites long.

Surprisingly, the TASO team converged on the same issue and added \texttt{Transpose(X) -> X} as a lemma when performing SMT verification for their rewrites. We adopt the same solution and enhance the axiom set with the lemmas provided by TASO.

Interestingly, this suggests that the ``distance'' in rewrite space and the ``distance'' in SMT proof space are closely correlated.
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\end{document}
